Tropical Mathematics and the Lambda Calculus

### 2023-01-24 section1.tex
commit

In recent years, more and more interest in the programming language community has been directed towards the study of \emph{quantitative} properties of programs like e.g.~computing the number of computation steps, or convergence probabilities, 
as opposed to purely \emph{qualitative} properties like e.g.~termination or program equivalence. 
Notably, a significant effort has been made to extend, or adapt, well-established qualitative methods, like e.g.~type systems, relational logics or denotational semantics, to account for quantitative properties. We can mention, for example, 
intersection type systems aimed at capturing time or space resources \cite{decarvalho2018, Accattoli2022} or convergence probabilities \cite{Breuvart2018, PistoneLICS2022},  relational logics to account for probabilistic properties like e.g.~differential privacy \cite{Barthe_2012} or metric preservation \cite{Reed2010, dallago}, as well as the study of denotational models for 
probabilistic \cite{Ehrhard2011, Staton2017} or differential \cite{difflambda} extensions of the $\lambda$-calculus. 
The main reason to look for methods relying on (quantitative extensions of) type-theory or denotational semantics is that these approaches yield \emph{modular} and \emph{compositional} techniques, that is, allow one to deduce properties of complex programs from the properties of their constituent parts.

@@

On the one hand one there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function (which is rarely the case), it makes more sense to ask   whether they compute functions which do not differ \emph{too much}. This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.

@@

A natural question is thus whether these two apparently unrelated ways of interpreting linearity and duplication can be somehow reconciled. At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
in metric models $n$ times duplication results in a \emph{linear} (hence Lipschitz) function $n\cdot x$, while in differential models this results in a \emph{polynomial} function $x^{n}$, hence not Lipschitz. The fundamental motivation of this work is then the observation that 
this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics, where, as we'll see, the monomial $x^{n}$ ``reads as'' the linear function $n\cdot x$.

@@

Coming back to our previous discussion on program semantics, tropical geometry might seem to provide precisely what we are looking for, as it turns the monomials $x^{n}$ into the Lipschitz functions $n\cdot x$.
At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail. Furthermore, connections between tropical linear algebra and metric spaces (in the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}), have also been observed \cite{Fuji}.

>>>>

In recent years, more and more interest in the programming language community has been directed towards the study of \emph{quantitative} properties of programs like computing the number of computation steps or convergence probabilities, 
as opposed to purely \emph{qualitative} properties like termination or program equivalence. 
Notably, a significant effort has been made to extend, or adapt, well-established qualitative methods, like type systems, relational logics or denotational semantics, to account for quantitative properties. We can mention, for example, 
intersection type systems aimed at capturing time or space resources \cite{decarvalho2018, Accattoli2022} or convergence probabilities \cite{Breuvart2018, PistoneLICS2022},  relational logics to account for probabilistic properties like e.g.~differential privacy \cite{Barthe_2012} or metric preservation \cite{Reed2010, dallago}, as well as the study of denotational models for 
probabilistic \cite{Ehrhard2011, Staton2017} or differential \cite{difflambda} extensions of the $\lambda$-calculus. 
The main reason to look for methods relying on (quantitative extensions of) type-theory or denotational semantics is that these approaches yield \emph{modular} and \emph{compositional} techniques, that is, allow one to deduce properties of complex programs from the properties of their constituent parts.

@@

On the one hand one there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask   whether they compute functions which do not differ \emph{too much}. This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.

@@

A natural question is thus whether these two apparently unrelated ways of interpreting linearity and duplication can be somehow reconciled. At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
in metric models $n$ times duplication results in a \emph{linear} (hence Lipschitz) function $n\cdot x$, while in differential models this results in a \emph{polynomial} function $x^{n}$, hence not Lipschitz. The fundamental motivation of this work is then the observation that 
this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics, where, as we'll see, the monomial $x^{n}$ precisely reads as the linear function $n\cdot x$.

@@

Coming back to our discussion on program semantics, tropical geometry might seem to provide precisely what look for, as it turns the monomials $x^{n}$ into the Lipschitz map  $n\cdot x$.
At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail. Furthermore, connections between tropical linear algebra and metric spaces have also been observed \cite{Fuji} within the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}.
However, a thorough investigation of the interpretation of the $\lambda$-calculus within tropical mathematics has not yet been undertaken.

### 2023-01-24 section3C.tex
commit

\begin{definition}
 The \emph{tropical differential operator} is the map $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho*\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (where a multiset $\nu \in !(X+X)$ is identified with a disjoint sum of $\mu,\rho\in !X$).
\end{definition}

\begin{remark}
For $t\in\HOM{\LREL}{!X}{Y}$, we have:
 $D^2 t\in\HOM{\LREL_!}{(X+X)+(X+X)}{Y}$, where $(D^2 t)_{(\rho\oplus\rho')\oplus(\nu\oplus\nu'),b}$ equals $t_{\nu*\nu'*\rho',b}$ if $\rho=\emptyset$ and $\#\rho'=1=\#\nu$; it equals $t_{\rho*\nu',b}$ if $\rho'=\emptyset=\nu$ and $\#\rho=1$; it equals $\infty$ otherwise.
\end{remark}

>>>>

\begin{definition}
 The \emph{tropical differential operator} is the map $D:\HOM{\LREL}{!X}{Y}\to \HOM{\LREL}{!(X+X)}{Y}$ defined as $(Dt)_{\mu\oplus\rho,b}=t_{\rho+\mu,b}$ if $\#\mu=1$ and as $\infty$ otherwise (where a multiset $\nu \in !(X+X)$ is identified with a disjoint sum of $\mu,\rho\in !X$).
\end{definition}

\begin{remark}
For $t\in\HOM{\LREL}{!X}{Y}$, we have:
 $D^2 t\in\HOM{\LREL_!}{(X+X)+(X+X)}{Y}$, where $(D^2 t)_{(\rho\oplus\rho')\oplus(\nu\oplus\nu'),b}$ equals $t_{\nu+\nu'+\rho',b}$ if $\rho=\emptyset$ and $\#\rho'=1=\#\nu$; it equals $t_{\rho+\nu',b}$ if $\rho'=\emptyset=\nu$ and $\#\rho=1$; it equals $\infty$ otherwise.
\end{remark}

### 2023-01-24 section4.tex
commit

We can now look at what happens with tLs, i.e.~when considering the full exponential $!$.
First, since a tropical polynomial can be represented as a tropical linear function from $\Lawv^{\C M_{\mathrm{deg}(\varphi)(X)}}$ to $\Lawv$, from Proposition \ref{prop:boundedlip} we deduce:
\begin{corollary}\label{prop:polylip}
Any tropical polynomials $\varphi:\Lawv^{X}\to\Lawv$ are $\mathrm{deg}(\varphi)$-Lipschitz continuous.
\end{corollary}

Using Corollary \ref{prop:polylip}, an immediate consequence of %the lines after Theorem~\ref{thm:modelsTaylor} is:
Corollary~\ref{cor:T(M)=M} is:

\begin{corollary}\label{thm:M=infLip}
 If $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is an $\inf$ of \emph{Lipschitz} functions.
\end{corollary}

% 
%
%The proposition above, together with Theorem \ref{theorem:fepsilon}, can be used to show that tLs with \emph{finitely many} variables are always \emph{locally} Lipschitz over $(0,\infty)^{X}$.

Moreover, as consequence of Theorem~\ref{theorem:fepsilon}, the tLs with \emph{finitely many} variables are always \emph{locally} Lipschitz on all $\BB R_{>0}$.
Actually, we can prove a more general statement, also covering the infinitary case.

@@

Finally, let us discuss the differential structure. The differential operator $D$ of $\LREL_{!}$ translates into a differential operator $D_{!}$ turing a tLs $f:\Lawv^{X}\to \Lawv^{Y}$ into a tLs $D_{!}f:\Lawv^{X}\times \Lawv^{X}\to \Lawv^{Y}$, linear in its first variable, and given by 
\begin{equation}
D_{!}f(x,y)_{b}=\inf_{a\in X, \mu\in !X}\left\{\matr f_{\mu*a}+x_{a}+\mu y\right\}
\end{equation}
One can check that, when $f$ is a tropical polynomial, $D_{!}f$ coincides with the standard tropical derivative (see e.g.~\cite{Grigoriev2017}).
Moreover, the Taylor formula \eqref{eq:taylorcat} yields a ``tropical'' Taylor formula for tLs of the form 
\begin{equation}
f(x)=\inf_{n}\left\{D_{!}^{(n)}(f)(!_{n}x,\infty)\right\}
\end{equation}
The following result shows that the distance between two tropical maps can be approximated using the terms appearing in their Taylor expansions:
\begin{proposition}
For all tLs $f,g: \Lawv^{X}\to \Lawv^{Y}$, and for all $n\in \BB N$, 
the functions $x\mapsto D_{!}^{(n)}(f)(!_{n}x,\infty)$ are $n$-Lipschitz. Moreover 
$\norm{ \matr f-\matr g}_{\infty}= \sup_{n} \norm{ {\delta^{(n)}f}- {\delta^{(n)}g}}_{\infty}$, 
where $\delta^{(n)}h$ indicates the matrix of $D_{!}^{(n)}h$.
%where $\delta^{(n)}f:( \Lawv^{X})^{n}\to \Lawv^{X}$ is the tropical linear function $\delta^{(n)}f(\B x_{1},\dots, \B x_{n})=
%(\Der^{(n)}f)(\B x_{1},\dots, \B x_{n}, \infty)$. 
\end{proposition}

>>>>

Observe that, for any tropical polynomial $\varphi:\Lawv^{X}\to \Lawv^{Y}$, the associated matrix has shape $!_{\mathrm{deg}(\varphi)}(X)\times Y\to \Lawv$ (as any monomial $ix+c_{i}$ yields a matrix entry on $!_{i}X\times Y$). Hence, using Proposition \ref{prop:boundedlip}, we have:
\begin{corollary}\label{prop:polylip}
For any tropical polynomial $\varphi:\Lawv^{X}\to\Lawv$, $\varphi$ is $\mathrm{deg}(\varphi)$-Lipschitz continuous.
\end{corollary}

Moreover, from Proposition \ref{prop:boundedlip}, we can also deduce the desired Lipschitz decomposition of the Taylor expansion (using Corollary~\ref{cor:T(M)=M}):

\begin{corollary}\label{thm:M=infLip}
 If $\Gamma \vdash_{\STLC}M:A$, then $\model M^!: \model\Gamma \to \model A$ is an $\inf$ of \emph{Lipschitz} functions (via its Taylor expansion \eqref{eq:taylorcat}).
\end{corollary}

Let us now look at what happens with tLs, i.e.~when considering the full exponential $!$.
As consequence of Theorem~\ref{theorem:fepsilon}, the tLs with \emph{finitely many} variables are always \emph{locally} Lipschitz on all $\BB R_{>0}$.
Actually, we can prove a more general statement, also covering the infinitary case.

@@

Finally, let us discuss the differential structure. The differential operator $D$ of $\LREL_{!}$ translates into a differential operator $D_{!}$ turing a tLs $f:\Lawv^{X}\to \Lawv^{Y}$ into a tLs $D_{!}f:\Lawv^{X}\times \Lawv^{X}\to \Lawv^{Y}$, linear in its first variable, and given by 
\begin{equation}
D_{!}f(x,y)_{b}=\inf_{a\in X, \mu\in !X}\left\{\matr f_{\mu+a}+x_{a}+\mu y\right\}
\end{equation}
One can check that, when $f$ is a tropical polynomial, $D_{!}f$ coincides with the standard tropical derivative (see e.g.~\cite{Grigoriev2017}).
Moreover, the Taylor formula \eqref{eq:taylorcat} yields a ``tropical'' Taylor formula for tLs of the form 
\begin{equation}
f(x)=\inf_{n}\left\{D_{!}^{(n)}(f)(!_{n}x,\infty)\right\}
\end{equation}
The following result shows that the distance between two tropical maps can be approximated using the terms appearing in their Taylor expansions:
\begin{proposition}
For all tLs $f,g: \Lawv^{X}\to \Lawv^{Y}$, and for all $n\in \BB N$, 
the functions $x\mapsto D_{!}^{(n)}(f)(!_{n}x,\infty)$ are $n$-Lipschitz. Moreover 
$\norm{ \matr f-\matr g}_{\infty}= \sup_{n} \norm{ {\delta^{(n)}f}- {\delta^{(n)}g}}_{\infty}$, 
where $\delta^{(n)}h$ indicates the matrix of $D_{!}^{(n)}h$.
%where $\delta^{(n)}f:( \Lawv^{X})^{n}\to \Lawv^{X}$ is the tropical linear function $\delta^{(n)}f(\B x_{1},\dots, \B x_{n})=
%(\Der^{(n)}f)(\B x_{1},\dots, \B x_{n}, \infty)$. 
\end{proposition}
